
\renewcommand{\thepage}{f. Budget Justification - Page \arabic{page} of 1}

\required{Budget Justification}
\begin{center}
\emph{Maximum of 3 pages}
\end{center}

\subsection*{Personnel}

No salary is requested for the PI, Co-PIs, or any Senior Personnel.

\subsection*{Other Personnel}
%Funds are requested to support one graduate student for 9 months
%during the academic year for each year of the project. At Arizona State University,
%the current pay rate for doctoral students at 50\% FTE is \$ during the academic
%year. Included is the estimated annual salary increase of 4\%.
%Brief description of students’ effort on the project should be provided.

\subsection*{Fringe Benefits}
Fringe benefits are applied to personnel salaries using the university approved rates:
\begin{itemize}
\item Faculty - \% in FYs 2012, 2013, and 2014
\item Postdocs - \% in FYs 2012, 2013, and 2014
\item Graduate students - \% in FYs 2012, 2013, and 2014
\item Undergraduate students - \% in FYs 2012, 2013, and 2014
\item Staff - \% in FYs 2012, 2013, and 2014
\item Part time staff - \% in FYs 2012, 2013, and 2014
\end{itemize}•

%Benefits are based on UC Davis policy for each of the following positions.  Benefits for Co-PI Ross-Ibarra are budgeted at 30.3% in year one, increasing to 39.7% in year 5.  Benefits for the postdoc are budgeted at 20.1% for the first 4 months of year one, increasing to 23.1% in the last 8 months of year 5.  The technician’s benefits start at 44.2% for the beginning of year 1 and increase to 58.4% in year 5. Graduate student benefits are budgeted at a constant 1.3% across all 5 years. 

\subsection*{Equipment}

No equipment funds are requested.

\subsection*{Travel}

Travel for the PI and Co-PI Coop and one student the postdoc to 1 domestic conference each year is budgeted at \$3,000.  

Travel for Senior Personnel and members of their group to manage field experiments and phenotype is budgeted at \$12,000 each of the first 3 years. Travel for both Senior Personnel to 1 international conference each year is budgeted at \$3,000 per year.

\subsection*{Other Direct Costs}

 \paragraph{Materials and Supplies:}\\
\paragraph{Whole genome sequencing}:
The genomes of each of the four parental lines of our QTL mapping populations will be resequenced to a depth of 20-30X using 2 lanes of paired end 150bp reads on an Illumina HiSeq 2500. Current lane costs are approximately \$2,200 per lane, and library preparations costs are approximately \$100, for a total cost of \$18,000.

\paragraph{RNA sequencing}:
In total, RNA sequencing will be performed on 192 individuals (8 inbreds x 2 stages x 2 tissues x 2 environments x 3 replicates + 8 NILs x 2 genotypes x 2 stages x 2 environs x 3 replicates).  Cost to prepare RNA libraries in our lab are approximately \$100 per library, and sequencing costs for single-end 50bp reads at the UCD Genome Center are approximately \$1,000 per lane.  Multiplexing 12 barcodes per lane, this comes out to 32 lanes of sequence and a total cost of \$70,400.

\paragraph{Field fees:}
Fees for the field experiments in our highland and lowland field sites \ref{tab:locales} are approximately \$60,000 the first three years of the experiment to allow development of the mapping populations and two replicates of the phenotyping.  These fees include land rental and basic management (planting, watering, weeding, fertilizing), as well as station fees to hire manual labor for phenotyping.  These fees decrease to \$10,000 in the last two years of the proposal as subsequent field experiments including evaluation of NILs and RNA-seq lines, will be considerably smaller. Field fees total \$200,000 across the five years of the grant.

\paragraph{Graduate Student Tuition:}
Tuition for graduate students is charged to the project in proportion to the amount of effort the graduate student will work on the project. For a graduate student employed on the project for 9 academic months at 50\% FTE, the tuition charge is \$31,546 in FY 2015 to account for out-of-state tuition, \$$17,266 in FY 2016 and increasing 5\% each subsequent year., \$ in FY 2013, and \$ in FY 2014.

\paragraph{Publication Costs:}
In year two \$1,500 is requested for publication fees to an open access journal.  In subsequent years \$3,000 is requested annually.

\subsection*{Total Direct Costs}

\subsection*{Indirect Costs}
Indirect costs are calculated on Modified Total Direct Costs (Total Direct costs less graduate student fees and participant support and subaward funding beyond the first \$25,000) using F\&A rates approved by US Department of Health and Human Services. For this project, F\&A rates of 55.5\% were used from Jan. 1, 2105 through June 30, 2015, 56.5\% from July 1, 2015 through June 30, 2016, and 57\% from July 1, 2016 until the end of the project.



